---
title: "Predictive Modelling Assignment"
author: "Anushka Muley"
date: "2026-03-20"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the necessary libraries
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(CalibrationCurves)
library(gam)
library(MASS)
library(pROC)
library(e1071)
library(tree)
```

## Data Preparation

### Loading the data

```{r}
hcv_data <- read.csv("C:/Users/anush/OneDrive/Desktop/Predictive Modelling/hcv+data/hcvdat0.csv")
```

### Inspecting Data

```{r}
head(hcv_data)
str(hcv_data)
```
### Cleaning and Imputation

```{r}
# Removal of X column as it is unnecessary for prediction
hcv_data <- dplyr::select(hcv_data, -X)

# Checking missing values
colSums(is.na(hcv_data))

# Handling missing values by simple imputation
numeric_vars <- names(hcv_data)[sapply(hcv_data, is.numeric)]
for(col in numeric_vars){
  hcv_data[[col]][is.na(hcv_data[[col]])] <- median(hcv_data[[col]], na.rm = TRUE)
}
```

### Encoding Variables and Creating Binary Outcome

```{r}
# Converting variables into factor
hcv_data$Sex <- factor(hcv_data$Sex, levels = c("f", "m"))


# Creating Binary Outcome
hcv_data$Outcome <- ifelse(hcv_data$Category %in% c("0=Blood Donor", "0s=Suspect Blood Donor"), 0, 1)

hcv_data$Outcome <- factor(hcv_data$Outcome,
                           levels = c(0,1),
                           labels = c("Healthy","Disease"))

table(hcv_data$Outcome)
```
# Train/Test Split

```{r}
# Splitting data into test and train
set.seed(1)
n <- nrow(hcv_data)          
train_index <- sample(n, size = 0.7 * n)  
train_data <- hcv_data[train_index, ]   
test_data  <- hcv_data[-train_index, ]  
```

### Scaling

```{r}
# Scaling
numeric_vars <- names(train_data)[sapply(train_data, is.numeric)]

# Scale training data
scaled_train <- scale(train_data[, numeric_vars])

# Scale test data
scaled_test <- scale(test_data[, numeric_vars],
                     center = attr(scaled_train, "scaled:center"),
                     scale = attr(scaled_train, "scaled:scale"))


train_data[, numeric_vars] <- scaled_train
test_data[, numeric_vars]  <- scaled_test
```

### Brier Score
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}

# Numeric outcome vectors for evaluation
y_train_numeric <- ifelse(train_data$Outcome == "Disease", 1, 0)
y_test_numeric  <- ifelse(test_data$Outcome  == "Disease", 1, 0)
```

---

### Model 1: Logistic Regression Model
```{r}
# Fit logistic regression on training data
glm_fit <- glm(Outcome ~ . - Category, data = train_data, family = binomial)

# Predicted probabilities for test set
prob_glm <- predict(glm_fit, newdata = test_data, type = "response")
summary(prob_glm)

# Brier score
brier_score(y_test_numeric, prob_glm)

```

```{r}
# Confusion Matrix
hcv.pred <- rep("Healthy", nrow(test_data))
hcv.pred[prob_glm > 0.5] <- "Disease"
table(hcv.pred, test_data$Outcome)
mean(hcv.pred == test_data$Outcome)
```
```{r}
#ROC Curve and AUC
roc_obj.l <- roc(y_test_numeric, prob_glm)   
plot(roc_obj.l, type = "s")
auc(roc_obj.l)
```

```{r}
# Calibration plot in test data
valProbggplot(prob_glm, y_test_numeric)
```

---

### Model 2: Logistic Regression with Spline on Age (GAM)

```{r}
gam.hcv<-gam(Outcome ~s(Age)+ Category+Sex+ALB+ALP+ALT+AST+BIL+CHE+CHOL+CREA+GGT+PROT
, data =train_data, family = binomial)
plot(gam.hcv)
hcv.probs.gam<-predict(gam.hcv, newdata = test_data, type = "response") 
brier_score(y_test_numeric, hcv.probs.gam)
```

```{r}
# Confusion Matrix
hcv.pred.gam <- rep("Healthy", nrow(test_data))
hcv.pred.gam[hcv.probs.gam > 0.5] <- "Disease"
table(hcv.pred.gam, test_data$Outcome)
mean(hcv.pred.gam == test_data$Outcome)
```

```{r}
# ROC Curve and AUC
roc_obj_gam <- roc(y_test_numeric, hcv.probs.gam)
plot(roc_obj_gam, type = "s", main = "ROC — GAM")
auc(roc_obj_gam)
```

```{r}
valProbggplot(hcv.probs.gam, y_test_numeric)
```
---

### Model 3: Linear Discriminant Analysis

```{r}
lda.fit <- lda(Outcome ~ ., data = train_data)
lda.fit
plot(lda.fit)
```

```{r}
lda.pred <- predict(lda.fit, test_data)
lda.class <- lda.pred$class
hcv.probs.lda  <- lda_pred$posterior[, 2]
# Confusion Matrix
lda_numeric <- ifelse(lda.class == "Disease", 1, 0)
table(lda_numeric, y_test_numeric)
mean(lda_numeric == y_test_numeric)
```

```{r}
# Posterior Distribution Summary
sum(lda.pred$posterior[, 2] < 0.5)
sum(lda.pred$posterior[, 2] >= 0.5)
```

```{r}
# ROC curve and AUC
roc_obj_lda <- roc(y_test_numeric, prob_lda)
plot(roc_obj_lda, type = "s", col = "slateblue3", main = "ROC — LDA")
auc(roc_obj_lda)
brier_score(y_test_numeric, prob_lda)
```

```{r}
valProbggplot(hcv.probs.lda, y_test_numeric)
```
---

### Model 4: Naive Bayes

```{r}
train_data$Sex <- as.factor(train_data$Sex)
test_data$Sex  <- as.factor(test_data$Sex)

nb.fit <- naiveBayes(Outcome ~ ., data = train_data)
nb.fit
```

```{r}
nb.class <- predict(nb.fit, test_data)
# Confusion Matrix
table(nb.class, test_data$Outcome)
mean(nb.class == test_data$Outcome)
nb.preds <- predict(nb.fit, newdata = test_data, type = "raw")
```

```{r}
brier_score(y_test_numeric, nb.preds[,2])
```

```{r}
# ROC Curve and AUC
roc_obj.nb <- roc(y_test_numeric, nb.preds[,2])
plot(roc_obj.nb, type = "s", col = "tomato3")
auc(roc_obj.nb)
```

```{r}
#Calibration Curve
valProbggplot(nb.preds[,2], y_test_numeric)
```
---

### Model 5: Decision Tree

```{r}
# Removing Category 
hcv_data <- dplyr::select(hcv_data, -Category)
train_data <- dplyr::select(train_data, -Category)
test_data  <- dplyr::select(test_data, -Category)
tree_model <- tree(Outcome ~ ., data = train_data)
```

```{r}
tree_pred <- predict(tree_model, newdata = test_data, type = "class")
```

```{r}
# Confusion Matrix
table(tree_pred, test_data$Outcome)
mean(tree_pred == test_data$Outcome)
```
```{r}
tree_prob <- predict(tree_model, newdata = test_data)[,2]
```

```{r}
brier_score(y_test_numeric, tree_prob)
```
```{r}
# ROC Curve and AUC 
roc_obj.tree <- roc(test_data$Outcome, tree_prob)
plot(roc_obj.tree)
auc(roc_obj.tree)
```
---

### Model Comparison: Summary Table

```{r}
# Logistic Regression
logit_brier <- brier_score(y_test_numeric, prob_glm)
logit_miscl <- mean(hcv.pred != test_data$Outcome)
logit_auc   <- auc(roc_obj.l)

# GAM
gam_brier <- brier_score(y_test_numeric, hcv.probs.gam)
gam_miscl <- mean(hcv.pred.gam != test_data$Outcome)
gam_auc   <- auc(roc_obj_gam)

# LDA
lda_brier <- brier_score(y_test_numeric, hcv.probs.lda)
lda_miscl <- mean(lda.class != test_data$Outcome)
lda_auc   <- auc(roc_obj_lda)

# Naive Bayes
nb_brier <- brier_score(y_test_numeric, nb.preds[,2])
nb_miscl <- mean(nb.class != test_data$Outcome)
nb_auc   <- auc(roc_obj.nb)

# Decision Tree
tree_brier <- brier_score(y_test_numeric, tree_prob)
tree_miscl <- mean(tree_pred != test_data$Outcome)
tree_auc   <- auc(roc_obj.tree)

# Create comparison table
perf <- data.frame(
  Model = c("Logistic Regression", "GAM", "LDA", "Naive Bayes", "Decision Tree"),
  Brier_Score = round(c(logit_brier, gam_brier, lda_brier, nb_brier, tree_brier), 4),
  Misclassification_Error = round(c(logit_miscl, gam_miscl, lda_miscl, nb_miscl, tree_miscl), 4),
  AUC = round(c(logit_auc, gam_auc, lda_auc, nb_auc, tree_auc), 4)
)

perf
```


### Model Comparison: ROC Curves

```{r}
plot(roc_obj.l, col = "red", main = "ROC curves")
lines(roc_obj_gam, col = "blue")
lines(roc_obj_lda, col = "green")
lines(roc_obj.nb, col = "purple")
lines(roc_obj.tree, col = "orange")
legend("bottomright", legend=c("Logistic","GAM","LDA","Naive Bayes","Tree"),
       col=c("red","blue","green","purple","orange"), lwd=2)
```

