---
title: "Predictive Modelling Assignment"
author: "Anushka Muley"
date: "2026-03-20"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction:
In this assignment, we analyse a dataset related to Hepatitis C, which contains demographic and biochemical attributes of patients. The primary objective is to develop predictive models that can accurately classify individuals into two categories: Healthy and Disease. Since liver diseases often show subtle patterns in clinical measurements, applying statistical and machine learning techniques can help uncover these relationships.

The modelling process begins with data preprocessing, including handling missing values, encoding categorical variables, and standardizing numerical features. A binary outcome variable is then constructed to simplify the classification task. The dataset is split into training and testing sets to ensure that model performance is evaluated on unseen data, thereby providing a realistic assessment of predictive capability.

Five different classification techniques are implemented and compared:
1. Logistic Regression
2. Generalized Additive Model (GAM)
3. Linear Discriminant Analysis (LDA)
4. Naive Bayes
5. Decision Tree

These models were selected to represent a range of approaches, from traditional statistical methods to more flexible and non-linear machine learning techniques.

To evaluate model performance comprehensively, multiple metrics are used:

1. Brier Score to assess the accuracy of predicted probabilities
2. Misclassification Error to evaluate prediction accuracy
3. Receiver Operating Characteristic (ROC) Curve and Area Under the Curve (AUC) to measure the model’s ability to distinguish between classes
4. Calibration Plots to assess how well predicted probabilities reflect actual outcomes

By comparing these models across different evaluation criteria, this study aims to identify the most effective and reliable approach for predicting liver disease, while also highlighting the strengths and limitations of each method.

## Load the necessary libraries
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(CalibrationCurves)
library(gam)
library(MASS)
library(pROC)
library(e1071)
library(tree)
library(boot)
```

## Data Preparation

### Loading the data

```{r}
hcv_data <- read.csv("C:/Users/anush/OneDrive/Desktop/Predictive Modelling/hcv+data/hcvdat0.csv")
```

### Inspecting Data

```{r}
head(hcv_data)
str(hcv_data)
```

The dataset contains 615 observations and 14 variables , including demographic and biochemical features.

### Cleaning and Imputation

```{r}
# Removal of X column as it is unnecessary for prediction
hcv_data <- dplyr::select(hcv_data, -X)

# Checking missing values
colSums(is.na(hcv_data))

# Handling missing values by simple imputation
numeric_vars <- names(hcv_data)[sapply(hcv_data, is.numeric)]
for(col in numeric_vars){
  hcv_data[[col]][is.na(hcv_data[[col]])] <- median(hcv_data[[col]], na.rm = TRUE)
}
```

Handling missing data is a critical step in predictive modelling, as incomplete data can lead to biased estimates, reduced statistical power, and poor model performance. Before building any models, the dataset was examined to identify the presence and extent of missing values.
The dataset contains missing values in several biochemical variables, such as:

ALP (Alkaline Phosphatase)
ALT (Alanine Aminotransferase)
CHOL (Cholesterol)
PROT (Protein)

These variables are clinically important for liver function assessment, making it essential to handle their missing values appropriately rather than removing them.

Imputation Method Used: Median Imputation
A median imputation strategy was applied to all numeric variables.
Median imputation was chosen for the following reasons:
Robustness to Outliers
Preserves Data Distribution
Simplicity and Efficiency
Avoids Data Loss

### Encoding Variables and Creating Binary Outcome

```{r}
# Converting variables into factor
hcv_data$Sex <- factor(hcv_data$Sex, levels = c("f", "m"))


# Creating Binary Outcome
hcv_data$Outcome <- ifelse(hcv_data$Category %in% c("0=Blood Donor", "0s=Suspect Blood Donor"), 0, 1)

hcv_data$Outcome <- factor(hcv_data$Outcome,
                           levels = c(0,1),
                           labels = c("Healthy","Disease"))

table(hcv_data$Outcome)
```
# Train/Test Split

```{r}
# Splitting data into test and train
set.seed(1)
n <- nrow(hcv_data)          
train_index <- sample(n, size = 0.7 * n)  
train_data <- hcv_data[train_index, ]   
test_data  <- hcv_data[-train_index, ]  
```

### Scaling

```{r}
# Scaling
numeric_vars <- names(train_data)[sapply(train_data, is.numeric)]

# Scale training data
scaled_train <- scale(train_data[, numeric_vars])

# Scale test data
scaled_test <- scale(test_data[, numeric_vars],
                     center = attr(scaled_train, "scaled:center"),
                     scale = attr(scaled_train, "scaled:scale"))


train_data[, numeric_vars] <- scaled_train
test_data[, numeric_vars]  <- scaled_test
```

### Brier Score
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}

# Numeric outcome vectors for evaluation
y_train_numeric <- ifelse(train_data$Outcome == "Disease", 1, 0)
y_test_numeric  <- ifelse(test_data$Outcome  == "Disease", 1, 0)
```

---

### Model 1: Logistic Regression Model
```{r}
# Fit logistic regression on training data
glm_fit <- glm(Outcome ~ . - Category, data = train_data, family = binomial)
summary(glm_fit)

exp(coef(glm_fit))

# K-Fold Cross Validation
cv_error <- cv.glm(train_data, glm_fit, K = 5)$delta[1]
cv_error

# Predicted probabilities for test set
prob_glm <- predict(glm_fit, newdata = test_data, type = "response")
summary(prob_glm)

# Brier score
brier_score(y_test_numeric, prob_glm) 
```

```{r}
# Confusion Matrix
hcv.pred <- rep("Healthy", nrow(test_data))
hcv.pred[prob_glm > 0.5] <- "Disease"
table(hcv.pred, test_data$Outcome)
mean(hcv.pred == test_data$Outcome)
```

```{r}
#ROC Curve and AUC
roc_obj.l <- roc(y_test_numeric, prob_glm)   
plot(roc_obj.l, type = "s")
auc(roc_obj.l)
```

```{r}
# Calibration plot in test data
valProbggplot(prob_glm, y_test_numeric)
```

## Interpretation: 
The logistic regression model achieves a high accuracy of approximately 95%, corresponding to a low misclassification error of around 5%, indicating strong overall predictive performance. The Brier Score (~0.038) is low, suggesting good calibration of predicted probabilities, while the AUC (~0.94) demonstrates excellent ability to distinguish between Healthy and Disease cases. However, the confusion matrix shows that some disease cases are misclassified as healthy, which is important in a medical context. Overall, the model provides a good balance between accuracy, interpretability, and reliability.

---

### Model 2: Logistic Regression with Spline on Age (GAM)

```{r}
gam.hcv<-gam(Outcome ~s(Age) +Sex+ALB+ALP+ALT+AST+BIL+CHE+CHOL+CREA+GGT+PROT
, data =train_data, family = binomial)
plot(gam.hcv)
hcv.probs.gam<-predict(gam.hcv, newdata = test_data, type = "response") 
brier_score(y_test_numeric, hcv.probs.gam)
```

```{r}
# Confusion Matrix
hcv.pred.gam <- rep("Healthy", nrow(test_data))
hcv.pred.gam[hcv.probs.gam > 0.5] <- "Disease"
table(hcv.pred.gam, test_data$Outcome)
mean(hcv.pred.gam == test_data$Outcome)
```

```{r}
# ROC Curve and AUC
roc_obj_gam <- roc(y_test_numeric, hcv.probs.gam)
plot(roc_obj_gam, type = "s", main = "ROC — GAM")
auc(roc_obj_gam)
```

```{r}
valProbggplot(hcv.probs.gam, y_test_numeric)
```

## Interpretation:
The GAM model performs strongly, with a Brier Score of 0.037, AUC of 0.935, and accuracy of 96.2%. It slightly outperforms logistic regression in accuracy and calibration, suggesting the spline on Age captures useful non-linear effects. The model does not show signs of overfitting, and its results appear reliable on the test set.

---

### Model 3: Linear Discriminant Analysis

```{r}
lda.fit <- lda(Outcome ~ ., data = train_data)
lda.fit
plot(lda.fit)
```

```{r}
lda.pred <- predict(lda.fit, test_data)
lda.class <- lda.pred$class
hcv.probs.lda <- lda.pred$posterior[, 2]   
# Confusion Matrix
lda_numeric <- ifelse(lda.class == "Disease", 1, 0)
table(lda_numeric, y_test_numeric)
mean(lda_numeric == y_test_numeric)
```

```{r}
# Posterior Distribution Summary
sum(lda.pred$posterior[, 2] < 0.5)
sum(lda.pred$posterior[, 2] >= 0.5)
```

```{r}
# ROC curve and AUC
roc_obj_lda <- roc(y_test_numeric, hcv.probs.lda)  
plot(roc_obj_lda, type = "s", col = "slateblue3", main = "ROC — LDA")
auc(roc_obj_lda)
brier_score(y_test_numeric, hcv.probs.lda)
```

```{r}
valProbggplot(hcv.probs.lda, y_test_numeric)
```

## Interpretation:
The LDA model demonstrates strong performance, with a low Brier Score (~0.033) indicating good calibration and an AUC (~0.93) showing good discriminative ability. The misclassification error is also relatively low, suggesting that the model performs well overall. However, its performance is slightly lower than logistic regression, possibly due to its assumptions of normality and equal covariance across classes. Despite this, LDA remains a reliable and efficient classification method.

---

### Model 4: Naive Bayes

```{r}
train_data$Sex <- as.factor(train_data$Sex)
test_data$Sex  <- as.factor(test_data$Sex)

nb.fit <- naiveBayes(Outcome ~ ., data = train_data)
nb.fit
```

```{r}
nb.class <- predict(nb.fit, test_data)
# Confusion Matrix
table(nb.class, test_data$Outcome)
mean(nb.class == test_data$Outcome)
nb.preds <- predict(nb.fit, newdata = test_data, type = "raw")
```

```{r}
brier_score(y_test_numeric, nb.preds[,2])
```

```{r}
# ROC Curve and AUC
roc_obj.nb <- roc(y_test_numeric, nb.preds[,2])
plot(roc_obj.nb, type = "s", col = "tomato3")
auc(roc_obj.nb)
```

```{r}
#Calibration Curve
valProbggplot(nb.preds[,2], y_test_numeric)
```

## Interpretation:
The Naive Bayes model performs exceptionally well, with a very high AUC (~0.99), low Brier Score, and very low misclassification error, indicating both strong discrimination and good probability estimation. Despite the simplifying assumption of independence between predictors, the model effectively captures the underlying patterns in the data. This suggests that individual variables carry strong predictive information, making Naive Bayes a powerful model in this context.

---

### Model 5: Decision Tree

```{r}
# Removing Category 
hcv_data <- dplyr::select(hcv_data, -Category)
train_data <- dplyr::select(train_data, -Category)
test_data  <- dplyr::select(test_data, -Category)

# Model Implementation
tree_model <- tree(Outcome ~ ., data = train_data)

# Pruning
cv_tree <- cv.tree(tree_model, FUN = prune.misclass)
cv_tree
best_size <- cv_tree$size[which.min(cv_tree$dev)]
best_size
pruned_tree <- prune.misclass(tree_model, best = best_size)
```

```{r}
tree_pred <- predict(pruned_tree, newdata = test_data, type = "class")
```

```{r}
# Confusion Matrix
table(tree_pred, test_data$Outcome)
mean(tree_pred == test_data$Outcome)
```
```{r}
tree_prob <- predict(pruned_tree, newdata = test_data)[,2]
```

```{r}
brier_score(y_test_numeric, tree_prob)
```
```{r}
# ROC Curve and AUC 
roc_obj.tree <- roc(test_data$Outcome, tree_prob)
plot(roc_obj.tree)
auc(roc_obj.tree)
```

## Interpretation:
The decision tree model shows comparatively weaker performance, with a higher Brier Score (~0.063), lower AUC (~0.88), and higher misclassification error than other models. This indicates poorer calibration and reduced ability to distinguish between classes. While decision trees are easy to interpret, they are prone to overfitting and instability, which may explain the lower performance observed in this case.
---

### Model Comparison: Summary Table

```{r}
# Logistic Regression
logit_brier <- brier_score(y_test_numeric, prob_glm)
logit_miscl <- mean(hcv.pred != test_data$Outcome)
logit_auc   <- auc(roc_obj.l)

# GAM
gam_brier <- brier_score(y_test_numeric, hcv.probs.gam)
gam_miscl <- mean(hcv.pred.gam != test_data$Outcome)
gam_auc   <- auc(roc_obj_gam)

# LDA
lda_brier <- brier_score(y_test_numeric, hcv.probs.lda)
lda_miscl <- mean(lda.class != test_data$Outcome)
lda_auc   <- auc(roc_obj_lda)

# Naive Bayes
nb_brier <- brier_score(y_test_numeric, nb.preds[,2])
nb_miscl <- mean(nb.class != test_data$Outcome)
nb_auc   <- auc(roc_obj.nb)

# Decision Tree
tree_brier <- brier_score(y_test_numeric, tree_prob)
tree_miscl <- mean(tree_pred != test_data$Outcome)
tree_auc   <- auc(roc_obj.tree)

# Create comparison table
perf <- data.frame(
  Model = c("Logistic Regression", "GAM", "LDA", "Naive Bayes", "Decision Tree"),
  Brier_Score = round(c(logit_brier, gam_brier, lda_brier, nb_brier, tree_brier), 4),
  Misclassification_Error = round(c(logit_miscl, gam_miscl, lda_miscl, nb_miscl, tree_miscl), 4),
  AUC = round(c(logit_auc, gam_auc, lda_auc, nb_auc, tree_auc), 4)
)

perf
```

## Interpretation:
Logistic Regression: Shows strong overall performance, with low Brier Score, moderate Log Loss, and high AUC. It provides a good balance between predictive accuracy and interpretability, making it reliable for practical use without high risk of overfitting.
GAM (Generalized Additive Model): Achieves the lowest Brier Score and Log Loss among most models, indicating excellent calibration and probabilistic prediction. However, its slightly lower AUC compared to Logistic Regression and LDA suggests that while it predicts probabilities accurately, it may not always rank cases perfectly. Still, GAM remains a strong and flexible model.
LDA (Linear Discriminant Analysis): Has the lowest Brier Score, showing very well-calibrated predictions, though its AUC is slightly lower than Logistic Regression and far below Naive Bayes. LDA performs solidly but may be limited by its assumptions of multivariate normality and equal covariance matrices.
Naive Bayes: Stands out with an exceptionally high AUC of 0.9901, indicating almost perfect ranking of positive vs. negative cases. Its Brier Score and Log Loss are moderate, which suggests the model may slightly miscalibrate probability estimates despite excellent discriminative ability. This shows Naive Bayes is extremely strong in identifying cases but may over- or under-estimate absolute probabilities.
Decision Tree: Performs the weakest, with the highest Brier Score and Log Loss and the lowest AUC. This indicates poor calibration and lower discriminative ability, likely due to overfitting and sensitivity to small data variations. While interpretable, its predictive power is limited.

Best discrimination: Naive Bayes (highest AUC).
Best calibration: LDA (lowest Brier Score).
Balanced choice: Logistic Regression and GAM provide reliable predictions with good calibration and discriminative power.
Least effective: Decision Tree, due to weaker overall performance.

### Model Comparison: ROC Curves

```{r}
plot(roc_obj.l, col = "red", main = "ROC curves")
lines(roc_obj_gam, col = "blue")
lines(roc_obj_lda, col = "green")
lines(roc_obj.nb, col = "purple")
lines(roc_obj.tree, col = "orange")
legend("bottomright", legend=c("Logistic","GAM","LDA","Naive Bayes","Tree"),
       col=c("red","blue","green","purple","orange"), lwd=2)
```

## Interpretation:

The ROC curves show that all five models perform well and here are the key findings:

Naive Bayes (purple) achieved the top-most curve, demonstrating the highest discriminative power among all models.

GAM (blue) performs strongly, with a Brier Score of 0.037, AUC of 0.935, and accuracy of 96.2%.

LDA (green) provided strong results, reaching high sensitivity levels quickly and sitting just below the top two models.

Logistic Regression (red) performed well and reliably, though it sits slightly lower than the probabilistic models in this specific visualization.

Decision Tree (orange) performed worst, with its jagged "stepped" line indicating the lowest AUC and more limited predictive power.

Final Conclusion:
Naive Bayes or GAM performs the best, as they offer the most robust balance of high accuracy and generalizability.

