---
title: "Predictive Modelling Assignment"
author: "Anushka Muley"
date: "2026-03-20"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction:
In this assignment, we analyse a dataset related to Hepatitis C, which contains demographic and biochemical attributes of patients. The primary objective is to develop predictive models that can accurately classify individuals into two categories: Healthy and Disease. Since liver diseases often show subtle patterns in clinical measurements, applying statistical and machine learning techniques can help uncover these relationships.

The modelling process begins with data preprocessing, including handling missing values, encoding categorical variables, and standardizing numerical features. A binary outcome variable is then constructed to simplify the classification task. The dataset is split into training and testing sets to ensure that model performance is evaluated on unseen data, thereby providing a realistic assessment of predictive capability.

Five different classification techniques are implemented and compared:
1. Logistic Regression
2. Generalized Additive Model (GAM)
3. Linear Discriminant Analysis (LDA)
4. Naive Bayes
5. Decision Tree

These models were selected to represent a range of approaches, from traditional statistical methods to more flexible and non-linear machine learning techniques.

To evaluate model performance comprehensively, multiple metrics are used:

1. Brier Score to assess the accuracy of predicted probabilities
2. Misclassification Error to evaluate prediction accuracy
3. Receiver Operating Characteristic (ROC) Curve and Area Under the Curve (AUC) to measure the model’s ability to distinguish between classes
4. Calibration Plots to assess how well predicted probabilities reflect actual outcomes

By comparing these models across different evaluation criteria, this study aims to identify the most effective and reliable approach for predicting liver disease, while also highlighting the strengths and limitations of each method.

## Load the necessary libraries
```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(CalibrationCurves)
library(gam)
library(MASS)
library(pROC)
library(e1071)
library(tree)
```

## Data Preparation

### Loading the data

```{r}
hcv_data <- read.csv("C:/Users/anush/OneDrive/Desktop/Predictive Modelling/hcv+data/hcvdat0.csv")
```

### Inspecting Data

```{r}
head(hcv_data)
str(hcv_data)
```

The dataset contains 615 observations and 14 variables , including demographic and biochemical features.

### Cleaning and Imputation

```{r}
# Removal of X column as it is unnecessary for prediction
hcv_data <- dplyr::select(hcv_data, -X)

# Checking missing values
colSums(is.na(hcv_data))

# Handling missing values by simple imputation
numeric_vars <- names(hcv_data)[sapply(hcv_data, is.numeric)]
for(col in numeric_vars){
  hcv_data[[col]][is.na(hcv_data[[col]])] <- median(hcv_data[[col]], na.rm = TRUE)
}
```

Handling missing data is a critical step in predictive modelling, as incomplete data can lead to biased estimates, reduced statistical power, and poor model performance. Before building any models, the dataset was examined to identify the presence and extent of missing values.
The dataset contains missing values in several biochemical variables, such as:

ALP (Alkaline Phosphatase)
ALT (Alanine Aminotransferase)
CHOL (Cholesterol)
PROT (Protein)

These variables are clinically important for liver function assessment, making it essential to handle their missing values appropriately rather than removing them.

Imputation Method Used: Median Imputation
A median imputation strategy was applied to all numeric variables.
Median imputation was chosen for the following reasons:
Robustness to Outliers
Preserves Data Distribution
Simplicity and Efficiency
Avoids Data Loss

### Encoding Variables and Creating Binary Outcome

```{r}
# Converting variables into factor
hcv_data$Sex <- factor(hcv_data$Sex, levels = c("f", "m"))


# Creating Binary Outcome
hcv_data$Outcome <- ifelse(hcv_data$Category %in% c("0=Blood Donor", "0s=Suspect Blood Donor"), 0, 1)

hcv_data$Outcome <- factor(hcv_data$Outcome,
                           levels = c(0,1),
                           labels = c("Healthy","Disease"))

table(hcv_data$Outcome)
```
# Train/Test Split

```{r}
# Splitting data into test and train
set.seed(1)
n <- nrow(hcv_data)          
train_index <- sample(n, size = 0.7 * n)  
train_data <- hcv_data[train_index, ]   
test_data  <- hcv_data[-train_index, ]  
```

### Scaling

```{r}
# Scaling
numeric_vars <- names(train_data)[sapply(train_data, is.numeric)]

# Scale training data
scaled_train <- scale(train_data[, numeric_vars])

# Scale test data
scaled_test <- scale(test_data[, numeric_vars],
                     center = attr(scaled_train, "scaled:center"),
                     scale = attr(scaled_train, "scaled:scale"))


train_data[, numeric_vars] <- scaled_train
test_data[, numeric_vars]  <- scaled_test
```

### Brier Score
```{r}
brier_score <- function(y, pred_prob) {
  mean((y - pred_prob)^2)
}

# Numeric outcome vectors for evaluation
y_train_numeric <- ifelse(train_data$Outcome == "Disease", 1, 0)
y_test_numeric  <- ifelse(test_data$Outcome  == "Disease", 1, 0)
```

---

### Model 1: Logistic Regression Model
```{r}
# Fit logistic regression on training data
glm_fit <- glm(Outcome ~ . - Category, data = train_data, family = binomial)

# Predicted probabilities for test set
prob_glm <- predict(glm_fit, newdata = test_data, type = "response")
summary(prob_glm)

# Brier score
brier_score(y_test_numeric, prob_glm)

```

```{r}
# Confusion Matrix
hcv.pred <- rep("Healthy", nrow(test_data))
hcv.pred[prob_glm > 0.5] <- "Disease"
table(hcv.pred, test_data$Outcome)
mean(hcv.pred == test_data$Outcome)
```

```{r}
#ROC Curve and AUC
roc_obj.l <- roc(y_test_numeric, prob_glm)   
plot(roc_obj.l, type = "s")
auc(roc_obj.l)
```

```{r}
# Calibration plot in test data
valProbggplot(prob_glm, y_test_numeric)
```

## Interpretation: 
The logistic regression model achieves a high accuracy of approximately 95%, corresponding to a low misclassification error of around 5%, indicating strong overall predictive performance. The Brier Score (~0.038) is low, suggesting good calibration of predicted probabilities, while the AUC (~0.94) demonstrates excellent ability to distinguish between Healthy and Disease cases. However, the confusion matrix shows that some disease cases are misclassified as healthy, which is important in a medical context. Overall, the model provides a good balance between accuracy, interpretability, and reliability.

---

### Model 2: Logistic Regression with Spline on Age (GAM)

```{r}
gam.hcv<-gam(Outcome ~s(Age)+ Category+Sex+ALB+ALP+ALT+AST+BIL+CHE+CHOL+CREA+GGT+PROT
, data =train_data, family = binomial)
plot(gam.hcv)
hcv.probs.gam<-predict(gam.hcv, newdata = test_data, type = "response") 
brier_score(y_test_numeric, hcv.probs.gam)
```

```{r}
# Confusion Matrix
hcv.pred.gam <- rep("Healthy", nrow(test_data))
hcv.pred.gam[hcv.probs.gam > 0.5] <- "Disease"
table(hcv.pred.gam, test_data$Outcome)
mean(hcv.pred.gam == test_data$Outcome)
```

```{r}
# ROC Curve and AUC
roc_obj_gam <- roc(y_test_numeric, hcv.probs.gam)
plot(roc_obj_gam, type = "s", main = "ROC — GAM")
auc(roc_obj_gam)
```

```{r}
valProbggplot(hcv.probs.gam, y_test_numeric)
```

## Interpretation:
The GAM model shows near-perfect performance, with an extremely low Brier Score (~0.000), an AUC of approximately 1.00, and almost zero misclassification error. This indicates that the model fits the data extremely well and can perfectly distinguish between classes. However, such perfect performance is likely a sign of overfitting, meaning the model may not generalize well to new data. While the flexibility of GAM allows it to capture non-linear relationships, its results should be interpreted with caution.

---

### Model 3: Linear Discriminant Analysis

```{r}
lda.fit <- lda(Outcome ~ ., data = train_data)
lda.fit
plot(lda.fit)
```

```{r}
lda.pred <- predict(lda.fit, test_data)
lda.class <- lda.pred$class
hcv.probs.lda <- lda.pred$posterior[, 2]   
# Confusion Matrix
lda_numeric <- ifelse(lda.class == "Disease", 1, 0)
table(lda_numeric, y_test_numeric)
mean(lda_numeric == y_test_numeric)
```

```{r}
# Posterior Distribution Summary
sum(lda.pred$posterior[, 2] < 0.5)
sum(lda.pred$posterior[, 2] >= 0.5)
```

```{r}
# ROC curve and AUC
roc_obj_lda <- roc(y_test_numeric, hcv.probs.lda)  
plot(roc_obj_lda, type = "s", col = "slateblue3", main = "ROC — LDA")
auc(roc_obj_lda)
brier_score(y_test_numeric, hcv.probs.lda)
```

```{r}
valProbggplot(hcv.probs.lda, y_test_numeric)
```

## Interpretation:
The LDA model demonstrates strong performance, with a low Brier Score (~0.033) indicating good calibration and an AUC (~0.93) showing good discriminative ability. The misclassification error is also relatively low, suggesting that the model performs well overall. However, its performance is slightly lower than logistic regression, possibly due to its assumptions of normality and equal covariance across classes. Despite this, LDA remains a reliable and efficient classification method.

---

### Model 4: Naive Bayes

```{r}
train_data$Sex <- as.factor(train_data$Sex)
test_data$Sex  <- as.factor(test_data$Sex)

nb.fit <- naiveBayes(Outcome ~ ., data = train_data)
nb.fit
```

```{r}
nb.class <- predict(nb.fit, test_data)
# Confusion Matrix
table(nb.class, test_data$Outcome)
mean(nb.class == test_data$Outcome)
nb.preds <- predict(nb.fit, newdata = test_data, type = "raw")
```

```{r}
brier_score(y_test_numeric, nb.preds[,2])
```

```{r}
# ROC Curve and AUC
roc_obj.nb <- roc(y_test_numeric, nb.preds[,2])
plot(roc_obj.nb, type = "s", col = "tomato3")
auc(roc_obj.nb)
```

```{r}
#Calibration Curve
valProbggplot(nb.preds[,2], y_test_numeric)
```

## Interpretation:
The Naive Bayes model performs exceptionally well, with a very high AUC (~0.99), low Brier Score, and very low misclassification error, indicating both strong discrimination and good probability estimation. Despite the simplifying assumption of independence between predictors, the model effectively captures the underlying patterns in the data. This suggests that individual variables carry strong predictive information, making Naive Bayes a powerful model in this context.

---

### Model 5: Decision Tree

```{r}
# Removing Category 
hcv_data <- dplyr::select(hcv_data, -Category)
train_data <- dplyr::select(train_data, -Category)
test_data  <- dplyr::select(test_data, -Category)
tree_model <- tree(Outcome ~ ., data = train_data)
```

```{r}
tree_pred <- predict(tree_model, newdata = test_data, type = "class")
```

```{r}
# Confusion Matrix
table(tree_pred, test_data$Outcome)
mean(tree_pred == test_data$Outcome)
```
```{r}
tree_prob <- predict(tree_model, newdata = test_data)[,2]
```

```{r}
brier_score(y_test_numeric, tree_prob)
```
```{r}
# ROC Curve and AUC 
roc_obj.tree <- roc(test_data$Outcome, tree_prob)
plot(roc_obj.tree)
auc(roc_obj.tree)
```

## Interpretation:
The decision tree model shows comparatively weaker performance, with a higher Brier Score (~0.063), lower AUC (~0.88), and higher misclassification error than other models. This indicates poorer calibration and reduced ability to distinguish between classes. While decision trees are easy to interpret, they are prone to overfitting and instability, which may explain the lower performance observed in this case.
---

### Model Comparison: Summary Table

```{r}
# Logistic Regression
logit_brier <- brier_score(y_test_numeric, prob_glm)
logit_miscl <- mean(hcv.pred != test_data$Outcome)
logit_auc   <- auc(roc_obj.l)

# GAM
gam_brier <- brier_score(y_test_numeric, hcv.probs.gam)
gam_miscl <- mean(hcv.pred.gam != test_data$Outcome)
gam_auc   <- auc(roc_obj_gam)

# LDA
lda_brier <- brier_score(y_test_numeric, hcv.probs.lda)
lda_miscl <- mean(lda.class != test_data$Outcome)
lda_auc   <- auc(roc_obj_lda)

# Naive Bayes
nb_brier <- brier_score(y_test_numeric, nb.preds[,2])
nb_miscl <- mean(nb.class != test_data$Outcome)
nb_auc   <- auc(roc_obj.nb)

# Decision Tree
tree_brier <- brier_score(y_test_numeric, tree_prob)
tree_miscl <- mean(tree_pred != test_data$Outcome)
tree_auc   <- auc(roc_obj.tree)

# Create comparison table
perf <- data.frame(
  Model = c("Logistic Regression", "GAM", "LDA", "Naive Bayes", "Decision Tree"),
  Brier_Score = round(c(logit_brier, gam_brier, lda_brier, nb_brier, tree_brier), 4),
  Misclassification_Error = round(c(logit_miscl, gam_miscl, lda_miscl, nb_miscl, tree_miscl), 4),
  AUC = round(c(logit_auc, gam_auc, lda_auc, nb_auc, tree_auc), 4)
)

perf
```

## Interpretation:
The comparison of models highlights that while several approaches perform well, their effectiveness varies in terms of reliability and practical usefulness. The GAM model stands out with perfect performance metrics; however, such results are unrealistic in real-world scenarios and strongly indicate overfitting. This suggests that although GAM is flexible and capable of capturing complex non-linear relationships, its predictions may not generalize well to new data.

The Naive Bayes model delivers exceptionally strong performance, with very high AUC and low Brier Score, showing that it can effectively distinguish between healthy and diseased individuals. Despite its simplifying assumption of independence between variables, it performs surprisingly well, indicating that the predictors individually carry strong signals.

The Logistic Regression model provides a very good balance between performance and interpretability. With high accuracy, strong AUC, and good calibration, it emerges as one of the most reliable models. Importantly, it avoids the overfitting issues seen in GAM while still maintaining strong predictive power, making it highly suitable for practical applications.

The LDA model also performs well, though slightly below logistic regression. Its performance is solid, but it relies on stricter assumptions about the data distribution, which may limit its flexibility in capturing complex relationships.

On the other hand, the Decision Tree model performs the weakest among all models. Although it is easy to interpret and useful for understanding decision rules, its lower AUC and higher Brier Score indicate poorer predictive performance. This is likely due to its tendency to overfit and its sensitivity to variations in the data.

In summary, while GAM appears to outperform all models, its results are likely misleading due to overfitting. Naive Bayes and Logistic Regression emerge as the most reliable models, offering strong predictive performance with better generalizability. The Decision Tree model, although interpretable, is the least effective for this dataset.

### Model Comparison: ROC Curves

```{r}
plot(roc_obj.l, col = "red", main = "ROC curves")
lines(roc_obj_gam, col = "blue")
lines(roc_obj_lda, col = "green")
lines(roc_obj.nb, col = "purple")
lines(roc_obj.tree, col = "orange")
legend("bottomright", legend=c("Logistic","GAM","LDA","Naive Bayes","Tree"),
       col=c("red","blue","green","purple","orange"), lwd=2)
```

## Interpretation:
The ROC curves show that all five models perform well and here are the key findings:
GAM achieved perfect performance, but likely overfitted
Naive Bayes performed exceptionally well, showing robustness
Logistic Regression provided strong and reliable results, balancing accuracy and interpretability
LDA performed well, slightly below logistic regression
Decision Tree performed worst, indicating limited predictive power
Final Conclusion:
Logistic Regression or Naive Bayes performs the best

